Operational Feature Selection with Gaussian Mixture Models for remote sensing images

### 2016-10-10 paper.tex
add fig for stopping crit + misc

    The computation of~\ref{eq:decision} requires the inversion of the
    covariance matrix and the  computation of the logarithm of the
    determinant.  The estimation of these terms suffers from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of parameters
    $\eta_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    a singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the high cost to collect ground-truth.

@@

        \subsubsection{Criterion function computation}
        \label{sec:update-crit}
        At iteration $k$, depending on the criterion, three terms have
        to  be computed:  the inverse  of the  covariance matrix,  the
        logarithm of the determinant of  the covariance matrix and the
        quadratic  term in  eq.~\ref{eq:decision}. However,  all these
        terms  have already  been computed  for iteration  $(k-1)$. By
        using  the  fact that  the  covariance  matrix is  a  positive
        definite     symmetric    matrix     and    block     matrices
        formulae~\cite[Chapter   9.2]{webb2003statistical},    it   is
        possible to factorize these terms at iteration $k$.

@@

        Using eq.~\ref{eq:cov} and~\ref{eq:cov:inv}, it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}[Update rule for the quadratical term]
        \label{eq:update-quad}
            \begin{align}
              (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once per selection step}}} \nonumber \\
                                                                                       &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}[Update rule for logdet]
        \label{eq:update-log}
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

        \begin{prop}[Decision function~\ref{eq:decision}]
            \begin{align}
            \label{eq:q-update}
                Q(\mathbf{x}) = &- \underbrace{(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber \\
                &- \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right) + 2 \log (\pi_c) + k \log(2 \pi)}_{\substack{\text{computed once per selection step}}} \nonumber \\
                &- \underbrace{\alpha \left( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] (\mathbf{x}^{(k)} - \boldsymbol{\mu}^{(k)}) \right)^2 - \log \alpha}_{\substack{\text{computed for each augmented set}}}.
            \end{align}
        \end{prop}

      \begin{prop}[Kullback-Leibler divergence~\ref{eq:skl}]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{ij} = \frac{1}{2} \biggl\{ \underbrace{\text{\normalfont Tr} \left((\boldsymbol{\Sigma}_i^{(k)})^{-1} \boldsymbol{\Sigma}_j^{(k)} + (\boldsymbol{\Sigma}_j^{(k)})^{-1} \boldsymbol{\Sigma}_i^{(k)}\right)}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_i}^t & \frac{1}{\alpha_i} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_j}^t & \frac{1}{\alpha_j} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2 - 2k}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t (\boldsymbol{\Sigma}_i^{(k-1)})^{-1}(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber\\
          & + \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t(\boldsymbol{\Sigma}_j^{(k-1)})^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \biggr\}.
        \end{align}
        with $(\boldsymbol{\Sigma}_i^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv};
      \end{prop}

      \begin{prop}[Jeffries-Matusita distance~\ref{eq:jm}]
      \begin{align}
        \label{eq:jm-update}
            \text{B}_{ij} = &\underbrace{\frac{1}{4} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t ( \boldsymbol{\tilde{\Sigma}}^{(k-1)} )^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{2} \ln \left( \frac{|\boldsymbol{\tilde{\Sigma}}^{(k-1)}|}{\sqrt{|\boldsymbol{\Sigma}_i^{(k-1)}| |\boldsymbol{\Sigma}_j^{(k-1)}|}} \right)}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{4} \tilde{\alpha} ( \left[\begin{array}{cc} \mathbf{\tilde{v}}^t & \frac{1}{\tilde{\alpha}} \end{array}\right] (\boldsymbol{\mu}_{i}^{(k)} - \boldsymbol{\mu}_{j}^{(k)}) )^2 + \frac{1}{2} \ln \left( \frac{\tilde{\alpha}}{\sqrt{\alpha_i \alpha_j}} \right)}_{\substack{\text{computed for each augmented set}}},
      \end{align}
      where $\boldsymbol{\tilde{\Sigma}} = \boldsymbol{\Sigma}_i + \boldsymbol{\Sigma}_j$ and $\tilde{\alpha}$ and $\mathbf{\tilde{v}}$ are defined as $\alpha$ and $\mathbf{v}$ but using $\boldsymbol{\tilde{\Sigma}}$ instead of $\boldsymbol{\Sigma}$.
    \end{prop}

@@

    \begin{algorithm}
    \caption{Sequential floating forward features selection with updates\label{alg:sffs-update}}
    {\footnotesize
    \begin{algorithmic}[1]
    \REQUIRE $J,\text{maxVarNb}$
    \STATE $\Omega=\overbrace{(\emptyset,...,\emptyset)}^{maxVarNb}$
    \STATE $F=\text{\{all variables\}}$
    \STATE $k=0$
    \WHILE{$k \neq \text{maxVarNb}$}
    \STATE Diagonalize $\boldsymbol{\Sigma}^{(k-1)} = \mathbf{Q} \boldsymbol{\Lambda} \mathbf{Q}^t$
    \STATE {\bfseries for all} $\Lambda_{i,i}$ {\bfseries do} $\Lambda_{i,i} = \max (\text{EPS\_FLT},\Lambda_{i,i})$
    \STATE Precompute {\scriptsize $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$, $(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)}- \boldsymbol{\mu}^{(k-1)})$ and $\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)$} using Propositions \ref{eq:update-inv}, \ref{eq:update-quad} and \ref{eq:update-log}
    \FORALL{$f_i \in F$}
    \STATE Compute update constant $\alpha$
    \STATE $\alpha = \max (\text{EPS\_FLT},\alpha)$
    \STATE $R_i = J(\{\Omega_k + f_i\})$ using Equations \ref{eq:q-update}, \ref{eq:skl-update} or \ref{eq:jm-update}
    \ENDFOR
    \STATE $j=\text{arg} \max_{i} R_i$
    \STATE $k=k+1$
    \IF{$R_j \geq J(\Omega_k)$}
    \STATE $\Omega_k = \{\Omega_{k-1} + f_j\}$
    \STATE $\text{flag}=1$
    \WHILE{$k > 2 \text{ and } \text{flag}=1$}
    \STATE Diagonalize $\boldsymbol{\Sigma}^{(k-1)} = = \mathbf{Q} \boldsymbol{\Lambda} \mathbf{Q}^t$
    \STATE {\bfseries for all} $\Lambda_{i,i}$ {\bfseries do} $\Lambda_{i,i} = \max (\text{EPS\_FLT},\Lambda_{i,i})$
    \STATE Compute {\scriptsize $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$, $(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)}- \boldsymbol{\mu}^{(k-1)})$ and $\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)$} using Propositions \ref{eq:update-inv}, \ref{eq:update-quad} and \ref{eq:update-log}
    \FORALL{$f_i \in \Omega_k$}
    \STATE Compute update constant $\alpha$
    \STATE $\alpha = \max (\text{EPS\_FLT},\alpha)$
    \STATE $R_i = J(\{\Omega_k \setminus f_i\})$ using Equations \ref{eq:q-update}, \ref{eq:skl-update} or \ref{eq:jm-update}
    \ENDFOR
    \STATE $j=\text{arg} \max_{i} R_i$
    \IF{$R_j > J(\Omega_{k-1})$}
    \STATE $\Omega_{k-1} = \{\Omega_k \setminus f_j\}$
    \STATE $k=k-1$
    \ELSE
    \STATE $\text{flag}=0$
    \ENDIF
    \ENDWHILE
    \ENDIF
    \ENDWHILE
    \RETURN $\Omega_{\text{maxVarNb}}$
    \end{algorithmic}
    }
    \end{algorithm}

@@

    For SFS and  SFFS selection, the number of variables  to select is
    set to  30 for the  Aisa dataset and  60 for the  Potsdam dataset.
    After  the selection  procedure,  the optimal  number of  selected
    variable is selected  as follow. Rather than  selecting the number
    of variables corresponding to the  highest value oft he criterion,
    the number of  retained variables is set when  the criterion stops
    to increase significantly.  It is found by  computing the discrete
    derivative of the criteria. See  %%TODO: faire une figure avec une
                                    %%courbe réelle et montrer où l'on
                                    %%s'arrete et où est le max.

    % Then, a small processing is made to select the number of variables used for prediction. The evolution of the criterion function is retrieved from the model file and the discrete derivative of the criterion \emph{DJ} is computed and normalized by its maximum. The subset of variables used for classification $\Omega^{*}$ is the first subset for which this value is lower than a given threshold \emph{th}, equal  to 0.001\% in this experiment,
    % \begin{equation*}
    %     \Omega^{*} = \Omega_{\min \{k|DJ_k<th\}}
    % \end{equation*}
    % with $DJ_k = \frac{J(\Omega_k) - J(\Omega_{k-1})}{\max_k (J(\Omega_k) - J(\Omega_{k-1}))}$.

@@

    In term of computational time, the GMM classifiers are as expected very fast for classification and also for training, except when the criterion function is a classification rate. In this case, using JM distance as criterion and SFS as search strategy is the best choice in term of time efficiency. The good performance in time can be explained by the dimension reduction. Actually, the decision rule corresponding to Equation~\ref{eq:decision} has a complexity in $d^3$ where $d$ is the dimension. Thus, reducing $d$ induces a reduction of the classification time.

@@

\appendices
\section{Proof of update rules}
\label{app:proof-update}
    % \subsubsection{Proposition~\ref{eq:update-inv-back}}
    %     \begin{proof}
    %         \begin{align*}
    %             \mathbf{A} - \alpha \mathbf{v} \mathbf{v}^t
    %             &= (\boldsymbol{\Sigma}^{(k-1)})^{-1} + \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \\
    %             &~~~- \alpha (- \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}) (- \frac{1}{\alpha} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1})) \\
    %             &= (\boldsymbol{\Sigma}^{(k-1)})^{-1}
    %         \end{align*}
    %     \end{proof}

>>>>

    The computation of eq.~(\ref{eq:decision}) requires the inversion of the
    covariance matrix and the  computation of the logarithm of the
    determinant.  The estimation of these terms suffers from the curse
    of  dimensionality\cite{bouveyron2014model}, because  in practice,
    the  number of  parameters  $\eta_c$ to  estimate  for each  class
    increases quadratically with respect to the number of features, as
    illustrated in Figure~\ref{fig:nb-param}.  Hence, if the number of
    observation  $n_c$ is  small  compared to  the  number of parameters
    $\eta_c$, the  estimated covariance  matrix is badly  conditioned and
    thus the computation  of its inverse and its  determinant would be
    unstable.   The worst  situation  is $n_c<\eta_c$  which leads  to
    a singular covariance matrix.  Unfortunately, this situation happens
    regularly in remote sensing.   For instance in hyperspectral image
    classification,  very few  labeled samples  are usually  available
    because of the difficulty and the high cost to collect ground-truth.

@@

        \subsubsection{Criterion function computation}
        \label{sec:update-crit}
        At iteration $k$, depending on the criterion, three terms have
        to  be computed:  the inverse  of the  covariance matrix,  the
        logarithm of the determinant of  the covariance matrix and the
        quadratic  term in  eq.(~\ref{eq:decision}). However,  all these
        terms  have already  been computed  for iteration  $(k-1)$. By
        using  the  fact that  the  covariance  matrix is  a  positive
        definite     symmetric    matrix     and    block     matrices
        formulae~\cite[Chapter   9.2]{webb2003statistical},    it   is
        possible to factorize these terms at iteration $k$.

@@

        Using eq.~(\ref{eq:cov}) and~(\ref{eq:cov:inv}), it is possible to
        deduce  the   following  propositions  (proof  are   given  in
        Appendix~\ref{app:proof-update}).
        \begin{prop}[Update rule for the quadratical term]
        \label{eq:update-quad}
            \begin{align}
              (\mathbf{x}^{(k)})^t (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)} = &\underbrace{(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)}}_{\substack{\text{computed once per selection step}}} \nonumber \\
                                                                                       &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2}_{\substack{\text{computed for each augmented set}}}.
            \end{align}
        \end{prop}
        \begin{prop}[Update rule for logdet]
        \label{eq:update-log}
            \begin{equation}
                \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) = \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)}_{\substack{\text{computed once}\\ \text{per selection step}}} + \underbrace{\log \alpha}_{\substack{\text{computed for} \\ \text{each augmented set}}}.
            \end{equation}
        \end{prop}

@@

        \begin{prop}[Decision function~(\ref{eq:decision})]
            \begin{align}
            \label{eq:q-update}
                Q(\mathbf{x}) = &- \underbrace{(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber \\
                &- \underbrace{\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right) + 2 \log (\pi_c) + k \log(2 \pi)}_{\substack{\text{computed once per selection step}}} \nonumber \\
                &- \underbrace{\alpha \left( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] (\mathbf{x}^{(k)} - \boldsymbol{\mu}^{(k)}) \right)^2 - \log \alpha}_{\substack{\text{computed for each augmented set}}}.
            \end{align}
        \end{prop}

      \begin{prop}[Kullback-Leibler divergence~(\ref{eq:skl})]
        \begin{align}
        \label{eq:skl-update}
            & SKL_{ij} = \frac{1}{2} \biggl\{ \underbrace{\text{\normalfont Tr} \left((\boldsymbol{\Sigma}_i^{(k)})^{-1} \boldsymbol{\Sigma}_j^{(k)} + (\boldsymbol{\Sigma}_j^{(k)})^{-1} \boldsymbol{\Sigma}_i^{(k)}\right)}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_i}^t & \frac{1}{\alpha_i} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{\alpha ( \left[\begin{array}{cc} \mathbf{v_j}^t & \frac{1}{\alpha_j} \end{array}\right] (\boldsymbol{\mu}_i^{(k)} - \boldsymbol{\mu}_j^{(k)}) )^2 - 2k}_{\substack{\text{computed for each augmented set}}} \nonumber \\
            &+ \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t (\boldsymbol{\Sigma}_i^{(k-1)})^{-1}(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber\\
          & + \underbrace{(\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t(\boldsymbol{\Sigma}_j^{(k-1)})^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \biggr\}.
        \end{align}
        with $(\boldsymbol{\Sigma}_i^{(k)})^{-1}$ computed with Proposition~\ref{eq:update-inv};
      \end{prop}

      \begin{prop}[Jeffries-Matusita distance~(\ref{eq:jm})]
      \begin{align}
        \label{eq:jm-update}
            \text{B}_{ij} = &\underbrace{\frac{1}{4} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})^t ( \boldsymbol{\tilde{\Sigma}}^{(k-1)} )^{-1} (\boldsymbol{\mu}_i^{(k-1)} - \boldsymbol{\mu}_j^{(k-1)})}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{2} \ln \left( \frac{|\boldsymbol{\tilde{\Sigma}}^{(k-1)}|}{\sqrt{|\boldsymbol{\Sigma}_i^{(k-1)}| |\boldsymbol{\Sigma}_j^{(k-1)}|}} \right)}_{\substack{\text{computed once per selection step}}} \nonumber \\
            &+ \underbrace{\frac{1}{4} \tilde{\alpha} ( \left[\begin{array}{cc} \mathbf{\tilde{v}}^t & \frac{1}{\tilde{\alpha}} \end{array}\right] (\boldsymbol{\mu}_{i}^{(k)} - \boldsymbol{\mu}_{j}^{(k)}) )^2 + \frac{1}{2} \ln \left( \frac{\tilde{\alpha}}{\sqrt{\alpha_i \alpha_j}} \right)}_{\substack{\text{computed for each augmented set}}},
      \end{align}
      where $\boldsymbol{\tilde{\Sigma}} = \boldsymbol{\Sigma}_i + \boldsymbol{\Sigma}_j$ and $\tilde{\alpha}$ and $\mathbf{\tilde{v}}$ are defined as $\alpha$ and $\mathbf{v}$ but using $\boldsymbol{\tilde{\Sigma}}$ instead of $\boldsymbol{\Sigma}$.
    \end{prop}

@@

    \begin{algorithm}
    \caption{Sequential floating forward features selection with updates\label{alg:sffs-update}}
    {\footnotesize
    \begin{algorithmic}[1]
    \REQUIRE $J,\text{maxVarNb}$
    \STATE $\Omega=\overbrace{(\emptyset,...,\emptyset)}^{maxVarNb}$
    \STATE $F=\text{\{all variables\}}$
    \STATE $k=0$
    \WHILE{$k \neq \text{maxVarNb}$}
    \STATE Diagonalize $\boldsymbol{\Sigma}^{(k-1)} = \mathbf{Q} \boldsymbol{\Lambda} \mathbf{Q}^t$
    \STATE {\bfseries for all} $\Lambda_{i,i}$ {\bfseries do} $\Lambda_{i,i} = \max (\text{EPS\_FLT},\Lambda_{i,i})$
    \STATE Precompute {\scriptsize $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$, $(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)}- \boldsymbol{\mu}^{(k-1)})$ and $\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)$} using Propositions (\ref{eq:update-inv}), (\ref{eq:update-quad}) and (\ref{eq:update-log})
    \FORALL{$f_i \in F$}
    \STATE Compute update constant $\alpha$
    \STATE $\alpha = \max (\text{EPS\_FLT},\alpha)$
    \STATE $R_i = J(\{\Omega_k + f_i\})$ using Equations (\ref{eq:q-update}), (\ref{eq:skl-update}) or (\ref{eq:jm-update})
    \ENDFOR
    \STATE $j=\text{arg} \max_{i} R_i$
    \STATE $k=k+1$
    \IF{$R_j \geq J(\Omega_k)$}
    \STATE $\Omega_k = \{\Omega_{k-1} + f_j\}$
    \STATE $\text{flag}=1$
    \WHILE{$k > 2 \text{ and } \text{flag}=1$}
    \STATE Diagonalize $\boldsymbol{\Sigma}^{(k-1)} = = \mathbf{Q} \boldsymbol{\Lambda} \mathbf{Q}^t$
    \STATE {\bfseries for all} $\Lambda_{i,i}$ {\bfseries do} $\Lambda_{i,i} = \max (\text{EPS\_FLT},\Lambda_{i,i})$
    \STATE Compute {\scriptsize $(\boldsymbol{\Sigma}^{(k-1)})^{-1}$, $(\mathbf{x}^{(k-1)} - \boldsymbol{\mu}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} (\mathbf{x}^{(k-1)}- \boldsymbol{\mu}^{(k-1)})$ and $\log \left(|\boldsymbol{\Sigma}^{(k-1)}|\right)$} using Propositions (\ref{eq:update-inv}), (\ref{eq:update-quad}) and (\ref{eq:update-log})
    \FORALL{$f_i \in \Omega_k$}
    \STATE Compute update constant $\alpha$
    \STATE $\alpha = \max (\text{EPS\_FLT},\alpha)$
    \STATE $R_i = J(\{\Omega_k \setminus f_i\})$ using Equations (\ref{eq:q-update}), (\ref{eq:skl-update}) or (\ref{eq:jm-update})
    \ENDFOR
    \STATE $j=\text{arg} \max_{i} R_i$
    \IF{$R_j > J(\Omega_{k-1})$}
    \STATE $\Omega_{k-1} = \{\Omega_k \setminus f_j\}$
    \STATE $k=k-1$
    \ELSE
    \STATE $\text{flag}=0$
    \ENDIF
    \ENDWHILE
    \ENDIF
    \ENDWHILE
    \RETURN $\Omega_{\text{maxVarNb}}$
    \end{algorithmic}
    }
    \end{algorithm}

@@

    For SFS and  SFFS selection, the number of variables  to select is
    set to  30 for the  Aisa dataset and  60 for the  Potsdam dataset.
    After  the selection  procedure,  the optimal  number of  selected
    variable is selected  as follow. Rather than  selecting the number
    of variables corresponding to the  highest value of the criterion,
    the number of  retained variables is set when  the criterion stops
    to increase significantly.  It is found by  computing the discrete
    derivative of the criteria. See \ref{fig:crit-evol}.

    \begin{figure}[!t]
        \centering
        \begin{tikzpicture}
            \begin{axis}[ymin=0.2,ymax=0.75,grid,axis x line=left,axis y line=left,xlabel={\# variables},ylabel={Cohen's kappa}]
                \addplot[thick,black] table[x=nb,y=crit] {criterion_1000spl.txt};
                \draw[thick,red] (axis cs:15,\pgfkeysvalueof{/pgfplots/ymin}) -- (axis cs:15,\pgfkeysvalueof{/pgfplots/ymax});
                \draw[thick,black] (axis cs:47,\pgfkeysvalueof{/pgfplots/ymin}) -- (axis cs:47,\pgfkeysvalueof{/pgfplots/ymax});
            \end{axis};
            % \begin{axis}[ymin=0.2,ymax=0.75,grid,axis x line=left,axis y line=left,xlabel={\# variables},ylabel={Cohen's kappa}]
            %     \addplot[thick,black] table[x=nb,y=crit] {criterion_50000spl.txt};
            %     \draw[thick,red] (axis cs:31,\pgfkeysvalueof{/pgfplots/ymin}) -- (axis cs:31,\pgfkeysvalueof{/pgfplots/ymax});
            %     \draw[thick,black] (axis cs:31,\pgfkeysvalueof{/pgfplots/ymin}) -- (axis cs:31,\pgfkeysvalueof{/pgfplots/ymax});
            % \end{axis};
        \end{tikzpicture}
        \caption{Criterion evolution in function of the number of selected variables. Red vertical line is the retained number of variables and black vertical line is the maximum of the criterion.\label{fig:crit-evol}}
    \end{figure}

    % Then, a small processing is made to select the number of variables used for prediction. The evolution of the criterion function is retrieved from the model file and the discrete derivative of the criterion \emph{DJ} is computed and normalized by its maximum. The subset of variables used for classification $\Omega^{*}$ is the first subset for which this value is lower than a given threshold \emph{th}, equal  to 0.1\% in this experiment,
    % \begin{equation*}
    %     \Omega^{*} = \Omega_{\min \{k|DJ_k<th\}}
    % \end{equation*}
    % with $DJ_k = \frac{J(\Omega_k) - J(\Omega_{k-1})}{\max_k (J(\Omega_k) - J(\Omega_{k-1}))}$.

@@

    In term of computational time, the GMM classifiers are as expected very fast for classification and also for training, except when the criterion function is a classification rate. In this case, using JM distance as criterion and SFS as search strategy is the best choice in term of time efficiency. The good performance in time can be explained by the dimension reduction. Actually, the decision rule corresponding to Equation~(\ref{eq:decision}) has a complexity in $d^3$ where $d$ is the dimension. Thus, reducing $d$ induces a reduction of the classification time.

@@

\appendices
\section{Proof of update rules}
\label{app:proof-update}
    % \subsubsection{Proposition~(\ref{eq:update-inv-back})}
    %     \begin{proof}
    %         \begin{align*}
    %             \mathbf{A} - \alpha \mathbf{v} \mathbf{v}^t
    %             &= (\boldsymbol{\Sigma}^{(k-1)})^{-1} + \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \\
    %             &~~~- \alpha (- \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}) (- \frac{1}{\alpha} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1})) \\
    %             &= (\boldsymbol{\Sigma}^{(k-1)})^{-1}
    %         \end{align*}
    %     \end{proof}

\begin{proof}[Proof of proposition~(\ref{eq:update-quad})]
    \begin{alignat*}{3}
    (&\mathbf{x}^{(k)})^t && (\boldsymbol{\Sigma}^{(k)})^{-1} \mathbf{x}^{(k)}
       = \left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t   & x_k \end{array}\right]
                                                            \left[\begin{array}{cc}
                                                                    \mathbf{A}   & \mathbf{v} \\
                                                                    \mathbf{v}^t & \frac{1}{\alpha}
                                                                  \end{array}\right]
                                                                                   \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= &&\left[\begin{array}{cc} (\mathbf{x}^{(k-1)})^t \mathbf{A} + x_k \mathbf{v}^t & (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} \end{array}\right]
                                                                                       \left[\begin{array}{c} \mathbf{v} \\ x_k \end{array}\right] \\
     &= &&(\mathbf{x}^{(k-1)})^t \mathbf{A} \mathbf{x}^{(k-1)} + x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + (\mathbf{x}^{(k-1)})^t \mathbf{v} x_k + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \frac{1}{\alpha} (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u} \mathbf{u}^t (\boldsymbol{\Sigma}^{(k-1)})^{-1}) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t ((\boldsymbol{\Sigma}^{(k-1)})^{-1} + \alpha \mathbf{v} \mathbf{v}^t) \mathbf{x}^{(k-1)}\\
     & &&+ 2 x_k \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha} \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} \mathbf{v}^t \mathbf{x}^{(k-1)} \\
     & &&+ 2 \frac{x_k}{\alpha} \mathbf{v}^t \mathbf{x}^{(k-1)} + \frac{(x_k)^2}{\alpha^2}) \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( (\mathbf{x}^{(k-1)})^t \mathbf{v} + \frac{x_k}{\alpha} )^2 \\
     &= &&(\mathbf{x}^{(k-1)})^t (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{x}^{(k-1)} + \alpha ( \left[\begin{array}{cc} \mathbf{v}^t & \frac{1}{\alpha} \end{array}\right] \mathbf{x}^{(k)} )^2
    \end{alignat*}
\end{proof}

\begin{proof}[Proof of proposition~(\ref{eq:update-log})]
    \begin{align*}
        \log \left(|\boldsymbol{\Sigma}^{(k)}|\right) =
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{u}      \\
                \mathbf{u}^t          & \sigma_{kk} \\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{0}_{k-1,1}     \\
                \mathbf{u}^t          & 1 \\
            \end{array}\bigg]
            \bigg[\begin{array}{cc}
                \mathbf{I}_{k-1} & (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}      \\
                \mathbf{0}_{k-1,1}^t          & \sigma_{kk} - \mathbf{u}^t(\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}\\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left(\left|
            \bigg[\begin{array}{cc}
                \boldsymbol{\Sigma}^{(k-1)} & \mathbf{0}_{k-1,1}     \\
                \mathbf{u}^t          & 1 \\
            \end{array}\bigg]
        \right|\right)
        +\log \left(\left|
            \bigg[\begin{array}{cc}
                \mathbf{I}_{k-1} & (\boldsymbol{\Sigma}^{(k-1)})^{-1} \mathbf{u}      \\
                \mathbf{0}_{k-1,1}^t          & \alpha \\
            \end{array}\bigg]
        \right|\right) \\
        &\log \left( |\boldsymbol{\Sigma}^{(k-1)}| |1| \right) + \log \left( |\mathbf{I}_{k-1}| |\alpha| \right)\\
        &\log \left( |\boldsymbol{\Sigma}^{(k-1)}| \right) + \log ( \alpha )
    \end{align*}
\end{proof}
